# Proposed addition to vignettes/ontoProc.Rmd
# Insert this section after the existing "onto_plot2" demonstration blocks,
# before the "# Related tools" section.

# ============================================================
# BEGIN INSERTION
# ============================================================

## Adaptive visualizations with ggplot2 (`onto_plot3`)

The `onto_plot2` function produces publication-ready graphs using
`Rgraphviz`, but its label handling is limited: very long ontological
term names can overflow node boundaries.  The new `onto_plot3` function
addresses this by building on the `ggraph` and `tidygraph` packages,
which sit on top of ggplot2.

Key improvements over `onto_plot2`:

* **Automatic label wrapping** — term names are broken at word boundaries
  using `strwrap()` so no string ever exceeds the configurable `width`
  character limit.
* **Auto-sizing node boxes** — `geom_node_label` draws a proper
  rectangular border that expands to fit the wrapped text.
* **Depth-coloured nodes** — each node is shaded by its distance from the
  root set, giving an at-a-glance view of hierarchy depth.
* **ggplot2 extensibility** — the returned object is a `ggplot`, so any
  ggplot2 layer, scale, or theme can be applied afterward.

```{r onto_plot3_basic, message=FALSE}
if (requireNamespace("ggraph", quietly = TRUE) &&
    requireNamespace("tidygraph", quietly = TRUE)) {
  library(ontoProc)
  cl = getOnto("cellOnto")
  cl3k = c("CL:0000492", "CL:0001054", "CL:0000236", "CL:0000625",
     "CL:0000576", "CL:0000623", "CL:0000451", "CL:0000556")
  p = onto_plot3(cl, cl3k, width = 20)
}
```

Because `onto_plot3` returns a `ggplot` object, further customisation is
straightforward:

```{r onto_plot3_custom, message=FALSE, eval=FALSE}
if (requireNamespace("ggraph", quietly = TRUE) &&
    requireNamespace("tidygraph", quietly = TRUE)) {
  library(ggplot2)
  p + theme(legend.position = "bottom") +
    labs(title = "PBMC cell types — Cell Ontology 2023",
         fill = "Hierarchy depth")
}
```

For ontologies that contain especially long term definitions (such as the
AEO anatomical entity ontology used earlier), increasing `width` prevents
text overflow:

```{r onto_plot3_aeo, message=FALSE, eval=FALSE}
if (requireNamespace("ggraph", quietly = TRUE) &&
    requireNamespace("tidygraph", quietly = TRUE)) {
  aeo = owl2cache(url="http://purl.obolibrary.org/obo/aeo.owl")
  aeoinr = setup_entities2(aeo)
  set.seed(1234)
  onto_plot3(aeoinr, sample(grep("AEO", names(aeoinr$name), value=TRUE), 12),
             width = 25, label_size = 2.5)
}
```

The `layout` argument accepts any layout algorithm supported by `ggraph`
or `igraph`.  The default `"sugiyama"` layout is optimal for directed
acyclic graphs (DAGs) such as ontologies, as it minimises edge crossings
and arranges nodes top-to-bottom by hierarchy level.  For exploratory use
with larger sub-graphs, `layout = "fr"` (Fruchterman-Reingold) or
`layout = "kk"` (Kamada-Kawai) may give more spread-out results:

```{r onto_plot3_layout, message=FALSE, eval=FALSE}
if (requireNamespace("ggraph", quietly = TRUE) &&
    requireNamespace("tidygraph", quietly = TRUE)) {
  onto_plot3(cl, cl3k, layout = "kk")
}
```

# ============================================================
# END INSERTION
# ============================================================
